Viscous Erosion in Porous Media

### 2018-03-13 docs/methods/erosion_methods.tex

With the shear stress computed, we are now ready to numerically evolve the boundary of each body using the {\thL} framework. The main equations to be solved are~\eqref{eqn:Levo1}--\eqref{eq:xsurf}.  In Section~\ref{sec:thetaL} we introduced a curvature-penalizing term into the erosion law~\eqref{eqn:Vn}. This term suppresses high-frequency oscillations, but also introduces {\em stiffness} into evolution equation~\eqref{eqn:thetaevo1}. We will therefore use a combined exponential-integrator/Runge-Kutta method to solve system~\eqref{eqn:Levo1}--\eqref{eq:xsurf}. The exponential integrator will handle the stiff term in a numerically stable fashion, while the Runge-Kutta method will treat nonlinear terms explicitly. Our overall time-stepping method will be second order.

@@

We now consider the system of evolution equations given by~\eqref{eqn:Levo1}--\eqref{eq:xsurf}. For convenience, we introduce the following variables
\begin{align}
%\label{eq:M}
& \Mterm = - 2 \pi \mean{\thalpha \Vnsig}, \\
%\label{eq:N}
& \NLterm = \frac{2 \pi}{L} \left( \pderiv{\atausig}{\alpha} + \thalpha
\Vssig \right), \\
%\label{eq:elfun}
& \elfun = \frac{2 \pi}{L}  \mean{\atausig} \, .
\end{align}
Then,~\eqref{eqn:Levo1} and~\eqref{eqn:thetaevo1} can be compactly expressed as
\begin{align}
\label{eqn:Levo2}
& \tderiv{L}{t} = \Mterm, \\
\label{eqn:thetaevo2}
& \pderiv{\theta}{t} = \eps \elfun \ppd{\theta}{\alpha} + \NLterm.
\end{align}
The first term on the right-hand-side of~\eqref{eqn:thetaevo2} is the stiff term resulting from the curvature penalization; i.e.~a curvature-driven interface velocity is associated with diffusion of the tangent angle $\theta$. This term is multiplied by $\zeta$, which is spatially constant but depends on time (due to $L$ and $\mean{\atau}$). Meanwhile, $\NLterm$ represents all remaining terms in the evolution equation, including contributions from the shear-dependent erosion law and from the artificial tangential velocity, $\Vs$. Since the local shear stress depends on interactions between bodies, $\NLterm$ is both a nonlinear and nonlocal.

We will evolve $\theta(\alpha,t)$ in spectral space and so introduce the Fourier series for $\theta$ and $\NLterm$
\begin{align}
\label{thetaSeries}
 \theta(\alpha,t) &= \alpha + \FourierSum \thhat_k e^{ i k \alpha} \, , \\
 \NLterm(\alpha,t)  &= \FourierSum \widehat{\NLterm}_k e^{ i k \alpha} \, .
\end{align}
The careful reader will notice that in~\eqref{thetaSeries} we have decomposed $\theta(\alpha)$ into a linear component and a periodic component. In spectral space,~\eqref{eqn:thetaevo2} produces the system of ODEs
\begin{align*}
\tderiv{\thhat_k}{t} +  \eps k^2  \elfun \thhat_k = \widehat{\NLterm}_k,
\qquad k=-N_\iin/2,\ldots,N_\iin/2-1\, .
\end{align*}
These ODEs can be simplified by applying the integrating factor
\begin{align}
\label{eq:mu}
\mu_k = \exp \left( \eps k^2 \int \elfun \, dt \right) \, ,
\end{align}
which leads to the coupled system
\begin{align}
\label{eqn:Levo3}
& \tderiv{L}{t} = \Mterm, \\
\label{eqn:thetaevo3}
& \tderiv{}{t}\left( \mu_k \thhat_k \right) = \mu_k \widehat{\NLterm}_k \, ,
\qquad k = -N_\iin/2, \ldots, N_\iin/2 -1.
\end{align}
Note that the diffusive term from~\eqref{eqn:thetaevo2} does not appear explicitly in~\eqref{eqn:thetaevo3}, but rather is implied by the integrating factor~\eqref{eq:mu}; i.e.~we have applied an exponential integrator. We solve the resulting system~\eqref{eq:mu}--\eqref{eqn:thetaevo3} with an
explicit, second-order Runge-Kutta method (RK2). In particular, we use the {\em midpoint} RK2 method so as to maximize the smoothing effect of the curvature penalization.
\todo[inline]{Last sentence might need to be clarified.  Not sure about
using `so as to'}

@@

At the second stage of RK2, we discretize~\eqref{eq:mu}--\eqref{eqn:thetaevo3} as 
\begin{align}
\label{eq:mu1d}
& \mu_k^{(n+1)} = \mu_k^{(n)} \exp \left( \eps k^2 {\Dt} \,
\zeta^{(n+1/2)} \right), \\
\label{eq:L1d}
& \frac{L^{(n+1)} - L^{(n)}}{\Dt} = \Mterm^{(n+1/2)}, \\
\label{eq:th1d}
& \frac{ \mu_k^{(n+1)} \thhat_k^{(n+1)} - \mu_k^{(n)} \thhat_k^{(n)}}{\Dt} =
\mu_k^{(n+1/2)} \hat{\NLterm}_k^{(n+1/2)}.
\end{align}
Once again, we must eventually compute the geometry at step $n+1$, and so we insert~\eqref{eq:mu1d} into~\eqref{eq:th1d} and solve for $\theta^{(n+1)}$, giving
\begin{align}
\thhat_k^{(n+1)} =  \thhat_k^{(n)} \exp \left( - \epsilon k^2 \Dt \elfun^{(n+1/2)} \right) + \Dt \, \hat{\NLterm}_k^{(n+1/2)} \exp \left( - \frac{1}{2} \eps k^2 \Dt \left( 2 \elfun^{(n+1/2)} - \elfun^{(n)} \right) \right).
\end{align}
Once again, both exponential terms act to smooth the geometry. Finally, to track the reference point, $\mean{x}$, we simply apply the same RK2 method to~\eqref{eq:xsurf}.

>>>>

With the shear stress computed, we are now ready to numerically evolve the boundary of each body using the {\thL} framework. The main equations to be solved are~\eqref{eqn:Levo1}--\eqref{eq:xsurf}.  In Section~\ref{sec:thetaL} we introduced a curvature-penalizing term into the erosion law~\eqref{eqn:Vn}. This term suppresses high-frequency oscillations, but also introduces {\em stiffness} into evolution equation~\eqref{eqn:thetaevo1}. We will therefore use a combined exponential-integrator/Runge-Kutta method to solve system~\eqref{eqn:Levo1}--\eqref{eq:xsurf}. The exponential integrator will handle the stiff term in a numerically stable fashion, while the Runge-Kutta method will treat nonlinear terms explicitly. Our overall time-stepping method will be second-order accurate.

@@

We now consider the system of evolution equations given by~\eqref{eqn:Levo1}--\eqref{eq:xsurf}. For convenience, we introduce the following variables
\begin{align}
%\label{eq:M}
& \Mterm = - 2 \pi \mean{\thalpha \Vnsig}, \\
%\label{eq:N}
& \NLterm = \frac{2 \pi}{L} \left( \pderiv{\atausig}{\alpha} + \thalpha
\Vssig \right), \\
%\label{eq:elfun}
& \elfun = \frac{2 \pi}{L}  \mean{\atausig} \, .
\end{align}
Then,~\eqref{eqn:Levo1} and~\eqref{eqn:thetaevo1} can be compactly expressed as
\begin{align}
\label{eqn:Levo2}
& \tderiv{L}{t} = \Mterm, \\
\label{eqn:thetaevo2}
& \pderiv{\theta}{t} = \eps \elfun \ppd{\theta}{\alpha} + \NLterm.
\end{align}
The diffusive term on the right of~\eqref{eqn:thetaevo2} results from the curvature penalization and is the stiffest term present. This term is multiplied by $\zeta$, which is spatially constant but depends on time (due to $L$ and $\mean{\atau}$). Meanwhile, $\NLterm$, represents all remaining terms in the evolution equation, including contributions from the shear-dependent erosion law and from the artificial tangential velocity, $\Vs$. Since the local shear stress depends on interactions between bodies, $\NLterm$ is both a nonlinear and nonlocal.

We will evolve $\theta(\alpha,t)$ in spectral space and so introduce the Fourier series for $\theta$ and $\NLterm$
\begin{align}
\label{thetaSeries}
 \theta(\alpha,t) &= \alpha + \FourierSum \thhat_k e^{ i k \alpha} \, , \\
 \NLterm(\alpha,t)  &= \FourierSum \widehat{\NLterm}_k e^{ i k \alpha} \, .
\end{align}
The careful reader will notice that in~\eqref{thetaSeries} we have decomposed $\theta(\alpha)$ into a linear component and a periodic component. In spectral space,~\eqref{eqn:thetaevo2} produces the system of ODEs
\begin{align*}
\tderiv{\thhat_k}{t} +  \eps k^2  \elfun \thhat_k = \widehat{\NLterm}_k,
\qquad k=-N_\iin/2,\ldots,N_\iin/2-1\, .
\end{align*}
These ODEs can be simplified by applying the integrating factor
\begin{align}
\label{eq:mu}
\mu_k = \exp \left( \eps k^2 \int \elfun \, dt \right) \, ,
\end{align}
which leads to the coupled system
\begin{align}
\label{eqn:Levo3}
& \tderiv{L}{t} = \Mterm, \\
\label{eqn:thetaevo3}
& \tderiv{}{t}\left( \mu_k \thhat_k \right) = \mu_k \widehat{\NLterm}_k \, ,
\qquad k = -N_\iin/2, \ldots, N_\iin/2 -1.
\end{align}
Note that the diffusive term from~\eqref{eqn:thetaevo2} does not appear explicitly in~\eqref{eqn:thetaevo3}, but rather is implied by the integrating factor~\eqref{eq:mu}. This is the idea of an exponential integrator. We solve the resulting system~\eqref{eq:mu}--\eqref{eqn:thetaevo3} with an
explicit, second-order Runge-Kutta method (RK2). In particular, we use the {\em midpoint} RK2 method so as to maximize the smoothing effect of the diffusive term.
\todo[inline]{Last sentence might need to be clarified.  Not sure about
using `so as to'}

@@

At the second stage of RK2, we discretize~\eqref{eq:mu}--\eqref{eqn:thetaevo3} as 
\begin{align}
\label{eq:mu1d}
& \mu_k^{(n+1)} = \mu_k^{(n)} \exp \left( \eps k^2 {\Dt} \,
\zeta^{(n+1/2)} \right), \\
\label{eq:L1d}
& \frac{L^{(n+1)} - L^{(n)}}{\Dt} = \Mterm^{(n+1/2)}, \\
\label{eq:th1d}
& \frac{ \mu_k^{(n+1)} \thhat_k^{(n+1)} - \mu_k^{(n)} \thhat_k^{(n)}}{\Dt} =
\mu_k^{(n+1/2)} \hat{\NLterm}_k^{(n+1/2)}.
\end{align}
We must eventually compute the geometry at step $n+1$, and so we insert~\eqref{eq:mu1d} into~\eqref{eq:th1d} and solve for $\theta^{(n+1)}$, giving
\begin{align}
\thhat_k^{(n+1)} =  \thhat_k^{(n)} \exp \left( - \epsilon k^2 \Dt \elfun^{(n+1/2)} \right) + \Dt \, \hat{\NLterm}_k^{(n+1/2)} \exp \left( - \frac{1}{2} \eps k^2 \Dt \left( 2 \elfun^{(n+1/2)} - \elfun^{(n)} \right) \right).
\end{align}
Once again, both exponential terms act to smooth the geometry. Finally, to track the reference point, $\mean{x}$, we simply apply the same RK2 method to~\eqref{eq:xsurf}.
